Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-06 manuscript.tex
Outline of conclusions

A tesseroid is a mass element defined in geocentric spherical
coordinates
(Figure~\ref{fig:tesseroid}).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~\ref{fig:tesseroid}).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational attraction,
and Marussi tensor, respectively,

@@

>>>>

A tesseroid is a mass element defined in geocentric spherical
coordinates
(Figure~\ref{fig:tesseroid}).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~\ref{fig:tesseroid}).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational acceleration,
and Marussi tensor, respectively,

@@

We have presented the open-source software package \textit{Tesseroids}.

It consists of command-line programs written in C.

The programs perform the forward modeling of gravitational fields in spherical
coordinates.

The fields are calculated from a mass model composed of spherical prisms, the
so-called tesseroids.

The volume integrals of the gravitational fields of a tesseroid are solved
numerically using the Gauss-Legendre Quadrature (GLQ).

The GLQ approximates the volume integral by a weighted sum of point mass
effects.

The error of the GLQ integration increases as the computation point gets closer
to the tesseroid.

The accuracy of the GLQ integration can be increased by using more point
masses.

Using more point masses also increases the computation time.

The tesseroid can be divided into smaller tesseroids as an alternative to
increasing the number of point masses in the GLQ.

The effect of each tesseroid is computed using the original number of point
masses.

This effectively increases the total number of point masses used.

We have implemented and improved upon an adaptive discretization algorithm to
achieve an optimal division of tesseroids.

Tesseroids are divided into more parts closer to the computation point.

This produces an optimal distribution of point masses needed to produce
accurate results.

Our implementation of the adaptive discretization uses a ``stack'' data
structure in place of the original recursive implementation.

The stack-based algorithm is computationally faster and allows more control
over errors when too many divisions are necessary (and the stack
``overflows'').

The criterion for dividing a tesseroid is based on the ratio the distance
(measured between the computation point and the geometric center of the
tesseroid) and the dimensions of the tesseroid.

The tesseroids is divided in two along a particular direction if this ratio
is larger than a parameter called the distance-size ratio ($D$).

Thus, by the end of the adaptive discretization algorithm, all tesseroids will
have dimensions smaller than $D$ times the distance to the computation point.

The value of $D$ indirectly controls the accuracy of the integration as well as
the computation time.

We performed an error analysis to determine the optimal value of $D$ to achieve
a target accuracy.

We used a spherical shell as a reference to determine the computation error of
our algorithm for different values of the distance-size ratio.

The spherical shell is convenient because there are analytical solutions for
its gravitational fields and it can be perfectly discretized into tesseroids.

We performed numerical experiments under different conditions.

We varied the size of the tesseroids used to discretize the shell, the height
of the computation points, and the location of the computation points (near the
pole or the equator).

We established a maximum allowed error of 0.1\% of the spherical shell
solutions.

The plots of the computation error per distance-size ratio reveal that the
values of $D$ required achieve 0.1\% error are
1 for the potential, 1.5 for the gravitational acceleration,
and 8 for the gravity gradients.

Previous assumptions were that accurate results are
guaranteed if the following condition is met:
the distance to the tesseroid is larger than the distance between
point masses (approximately the size of the tesseroid for a second-order GLQ).

This same condition was applied indiscriminately to both the gravitational
acceleration and the gravity gradients.

This condition is equivalent to using a distance-size ratio of 1.5.

Our results show that using $D=1$ produces an acceptable error of
approximately 0.1\% for the gravitational acceleration.

However, the same condition applied to the gravity gradients produces
an error of the order of $10^2\%$.

Therefore, the higher frequency gravity gradients require a finer
discretization of the tesseroids.

The behavior of the error for the gravity gradients varies with the computation
height.

The distance-size ratio required for 0.1\% error decreases with height.

This can be attributed to the higher sensitivity of gravity gradient to height
variations.

The integration kernel for gravity gradients has high frequency components
and thus requires more point masses for numerical integration.

We chose the value $D=8$ as a conservative default.

The values of the distance-size ratio determined in the error analysis were
incorporated as defaults in the software \textit{Tesseroids}.

The value of $D$ used can be controlled by a command-line argument.

In situations that require many tesseroid divisions in the adaptive
discretization, the stack used in the algorithm will overflow and further
divisions become impossible.

The current version of \textit{Tesseroids} (1.2) warns the user that
this situation occurred and proceeds with the GLQ integration without
division.

Future improvements to the algorithm include a better way to handle situations
that require many tesseroid divisions in the adaptive discretization.

An alternative would be to replace the tesseroid by an equivalent
right-rectangular prism and compute its effects instead.

Also, the computation time increases drastically as the computation point
gets closer to the tesseroid.

This effect can be prohibitive for computing the gravity gradients at
relatively low heights (e.g., for ground or airborne surveys).

Further investigation of different criteria for dividing the tesseroids could
yield better performance through a reduced number of divisions.
